Perturbing SU($n$)-symmetric interactions

### 2019-11-01 collective_sun/sun_perturbation.tex
some rearranging text

We consider an array of $N$ multilevel spins with non-uniform
SU($n$)-symmetric interactions that can be written in the form
\begin{align}
  H_0 = \sum_{p<q} h_{pq} \Pi_{pq},
  &&
  \Pi_{pq} \equiv \sum_{\mu,\nu} S_{\mu\nu}^{(p)} S_{\nu\mu}^{(q)},
  \label{eq:ints}
\end{align}
where $S_{\mu\nu}^{(p)}\equiv\op{\mu}{\nu}_p$ flips the state of spin
$p$ to $\ket\mu$ from $\ket\nu$; the operator $\Pi_{pq}$ permutes
spins $p$ and $q$; and the $h_{pq}$ in the Hamiltonian are all
negative: $h_{pq}<0$.

The ground-state manifold $\M_0$ of the interaction Hamiltonian $H_0$
consists of fully symmetric states that are simultaneous $+1$
eigenstates of all permutation operators $\Pi_{pq}$.  We will assume
throughout these notes that the interaction Hamiltonian $H_0$ is
gapped, i.e.~that there is a finite energy difference between $\M_0$
and its orthogonal complement.  The energy of fully symmetric states
$\ket\psi\in\M_0$ is
\begin{align}
  E_0 \equiv \sum_{p<q} h_{pq}
  = \f12 \sum_{p,q} h_{pq}
  = \f12 \sum_p h_p
  = \f12 N h,
\end{align}
where for convenience define coefficients $h_{pq}$ for all $p,q$ (as
opposed to only $p<q$) by enforcing $h_{pq}=h_{qp}$ and $h_{pp}=0$;
and
\begin{align}
  h_p \equiv \sum_q h_{pq},
  &&
  h \equiv \EE_p\sp{h_p} = \f1N \sum_p h_p.
\end{align}
We wish to determine the effective dynamics induced on the fully
symmetric manifold $\M_0$ by weak single- and two-body perturbations
of the form
\begin{align}
  \V_1 \equiv \sum_p v_p V_p = \v v\c\v V,
  &&
  \V_2 \equiv \sum_{p<q} w_{pq} X_p Y_q
  = \f12 \sum_{p,q} w_{pq} X_p Y_q
  = \f12 \v X \c\m w \c\v Y,
  \label{eq:perturbations}
\end{align}
where $V_p,X_p,Y_p$ are trace-zero operators on spin $p$;
$\v V,\v X,\v Y$ are respectively vectors of all $V_p,X_p,Y_p$; $\v v$
and $\m w$ are respectively a vector and matrix of all $v_p,w_{pq}$;
we enforce $w_{pq}=w_{qp}$ and $w_{pp}=0$; and we use a dot ($\c$) to
denote an inner product between matrices or vectors as appropriate.

The effective Hamiltonian $H_M$ induced on the ground-state manifold
$\M_0$ by an $M$-body perturbation $\V_M$ through second order in
perturbation theory is given by\cite{bravyi2011schrieffer,
  perlin2019effective}
\begin{align}
  H_M = H_M^{(1)} + H_M^{(2)},
  &&
  H_M^{(1)} = \P_0 \V_M \P_0,
  &&
  H_M^{(2)} = - \P_0 \V_M \E \V_M \P_0,
  &&
  \E \equiv \sum_{\Delta>0} \f{\P_\Delta}{\Delta},
\end{align}
where $\P_\Delta$ is a projector onto the eigenspace of the
interaction Hamiltonian $H_0$, with interaction energy $\Delta$
above that of fully symmetric manifold $\M_0$.

@@

To compute the effective Hamiltonian induced by the single-body
perturbation $\V_1$, we expand
\begin{align}
  H_1^{(1)} = \v v \c \P_0 \v V \P_0,
\end{align}
and use the permutational symmetry of the ground-state manifold to
simplify
\begin{align}
  H_1^{(1)} =  \sum_p v_p \P_0 V_0 \P_0 \simeq v V,
  &&
  v \equiv \EE_p\sp{v_p},
  &&
  V \equiv \sum_p V_p,
  \label{eq:H_1_1}
\end{align}
where $\simeq$ denotes equality up to a restriction to the
ground-state manifold $\M_0$.  In order to compute the second order
effective Hamiltonian $H_1^{(2)}$, we first choose an arbitrary fully
symmetric state $\ket\psi\in\M_0$ and expand (see Appendix
\ref{sec:H_V1_psi}):
\begin{align}
  H_0 \V_1 \ket\psi
  = E_0 \V_1 \ket\psi
  + \p{\v v\c\m h - \v v\oc\v h}\c \v V \ket\psi,
  \label{eq:H_V1_psi_start}
\end{align}
where $\m h$ is a matrix of all coefficients $h_{pq}$ in the
interaction Hamiltonian $H_0$ in \eqref{eq:ints}, and $\v v\oc\v h$ is
the element-wise (Hadamard) product of $\v v$ and $\v h$, i.e.~a
vector with components $v_kh_k$.  We thus find that the vector
$\V_1\ket\psi$ is an eigenvector of the interaction Hamiltonian $H_0$
if the coefficients $\v v$ satisfy
\begin{align}
  \m h \c \v v - \v h\oc\v v
  = \p{\m h - \diag\v h}\c\v v
  = \Delta\p{\v v} \v v,
  \label{eq:cond_1}
\end{align}
where $\diag\v h$ is a matrix with $\v h$ on the diagonal and zeroes
everywhere else; and $\Delta\p{\v v}$ is some constant that may depend
on $\v v$.  If the interaction Hamiltonian $H_0$ is translationally
invariant, then all $h_q=h$ and $\v h\oc\v v=h\v v$, so the condition
in \eqref{eq:cond_1} simplifies to
\begin{align}
  \m h \c\v v = \sp{\Delta\p{\v v}+h} \v v.
\end{align}
If we {\it construct} a perturbation $\V_1\p{\v v_\Delta}$ with
coefficients $\v v_\Delta$ that satisfy \eqref{eq:cond_1} with
eigenvalue $\Delta$, then
\begin{align}
  H_0 \V_1\p{\v v_\Delta} \ket\psi
  = \sp{E_0 + \Delta} \V_1\p{\v v_\Delta} \ket\psi.
  \label{eq:H_V1_psi}
\end{align}
Interestingly, the energy of the (unnormalized) state
$\V_1\p{\v v}\ket\psi$ depends only on the coefficients $\v v$, and is
entirely independent of the state $\ket\psi$ or choice of single-spin
trace-zero operator used to build $\V_1$ in \eqref{eq:perturbations}.
Finding operators $\V_1\p{\v v}$ that generate eigenvectors of the
interaction Hamiltonian $H_0$ when they act on fully symmetric states
$\ket\psi$ thus reduces to finding eigenvectors $\v v$ of the
interaction matrix $\m h-\diag\v h$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Two-body product at first order}
\label{sec:PXYP}

>>>>

We consider an array of $N$ multilevel spins with non-uniform
SU($n$)-symmetric interactions that can be written in the form
\begin{align}
  H_0 = \sum_{p<q} h_{pq} \Pi_{pq}
  = \f12 \sum_{p,q} \Pi_{pq},
  &&
  \Pi_{pq} \equiv \sum_{\mu,\nu} S_{\mu\nu}^{(p)} S_{\nu\mu}^{(q)},
  \label{eq:ints}
\end{align}
where the coefficients $h_{pq}$ in the Hamiltonian are all negative,
with $h_{pq}=h_{qp}$ and $h_{pp}=0$;
$S_{\mu\nu}^{(p)}\equiv\op{\mu}{\nu}_p$ flips the state of spin $p$ to
$\ket\mu$ from $\ket\nu$; and the operator $\Pi_{pq}$ permutes spins
$p$ and $q$.  The permutation operator $\Pi_{pq}$ is a multilevel
generalization of the SU(2)-symmetric spin interaction
$\m S_p\c\m S_q$.

The ground-state manifold $\M_0$ of the interaction Hamiltonian $H_0$
consists of fully symmetric states that are simultaneous $+1$
eigenstates of all permutation operators $\Pi_{pq}$.  We wish to
determine the effective dynamics induced on the fully symmetric
manifold $\M_0$ by weak single- and two-body perturbations of the form
\begin{align}
  \V_1 \equiv \sum_p v_p V_p = \v v\c\v V,
  &&
  \V_2 \equiv \sum_{p<q} w_{pq} X_p Y_q
  = \f12 \sum_{p,q} X_p w_{pq} Y_q
  = \f12 \v X \c\m w \c\v Y,
  \label{eq:perturbations}
\end{align}
where $V_p,X_p,Y_p$ are trace-zero operators on spin $p$;
$\v V,\v X,\v Y$ are respectively vectors of all $V_p,X_p,Y_p$; $\v v$
and $\m w$ are respectively a vector and matrix of all $v_p,w_{pq}$;
we enforce $w_{pq}=w_{qp}$ and $w_{pp}=0$; and we use a dot ($\c$) to
denote an inner product between vectors and matrices as appropriate.

We will assume throughout these notes that the interaction Hamiltonian
$H_0$ is gapped, and that the interaction energy difference between
$\M_0$ and its orthogonal complement is much larger than than any
additional couplings that we introduce.  The energy of fully symmetric
states $\ket\psi\in\M_0$ is
\begin{align}
  E_0 = \f12 \sum_{p,q} h_{pq}
  = \f12 \sum_p h_p
  = \f12 N h,
\end{align}
where for later convenience we define the net interaction energy $h_p$
associated with spin $p$, the mean $h$ of all such energies:
\begin{align}
  h_p \equiv \sum_q h_{pq},
  &&
  h \equiv \EE_p\sp{h_p} = \f1N \sum_p h_p.
\end{align}
The effective Hamiltonian $H_M$ induced on the ground-state manifold
$\M_0$ by an $M$-body perturbation $\V_M$ through second order in
perturbation theory is given by\cite{bravyi2011schrieffer,
  perlin2019effective}
\begin{align}
  H_M = H_M^{(1)} + H_M^{(2)},
  &&
  H_M^{(1)} = \P_0 \V_M \P_0,
  &&
  H_M^{(2)} = - \P_0 \V_M \E \V_M \P_0,
  &&
  \E \equiv \sum_{\Delta>0} \f{\P_\Delta}{\Delta},
\end{align}
where $\P_\Delta$ is a projector onto the eigenspace of the
interaction Hamiltonian $H_0$, with interaction energy $\Delta$ above
that of fully symmetric manifold $\M_0$; that is, the energy of states
in the image of $\P_\Delta$ is $E_0+\Delta$.

@@

To compute the effective Hamiltonian induced by the single-body
perturbation $\V_1$, we expand
\begin{align}
  H_1^{(1)} = \v v \c \P_0 \v V \P_0,
\end{align}
and use the permutational symmetry of the ground-state manifold to
simplify
\begin{align}
  H_1^{(1)} =  \sum_p v_p \P_0 V_0 \P_0 \simeq v V,
  &&
  v \equiv \EE_p\sp{v_p},
  &&
  V \equiv \sum_p V_p,
  \label{eq:H_1_1}
\end{align}
where $\simeq$ denotes equality up to a restriction to the
ground-state manifold $\M_0$.  In order to compute the second order
effective Hamiltonian $H_1^{(2)}$, we first choose an arbitrary fully
symmetric state $\ket\psi\in\M_0$ and expand (see Appendix
\ref{sec:H_V1_psi}):
\begin{align}
  H_0 \V_1 \ket\psi
  = E_0 \V_1 \ket\psi
  + \p{\v v\c\m h - \v v\oc\v h}\c \v V \ket\psi,
  \label{eq:H_V1_psi_start}
\end{align}
where $\m h$ is a matrix of all coefficients $h_{pq}$ of the
interaction Hamiltonian $H_0$ in \eqref{eq:ints}, and $\v v\oc\v h$ is
the element-wise (Hadamard) product of $\v v$ and $\v h$, i.e.~a
vector with components $v_kh_k$.  We thus find that the vector
$\V_1\ket\psi$ is an eigenvector of the interaction Hamiltonian $H_0$
if the coefficients $\v v$ satisfy
\begin{align}
  \m h \c \v v - \v h\oc\v v
  = \p{\m h - \diag\v h}\c\v v
  = \Delta\p{\v v} \v v,
  \label{eq:cond_1}
\end{align}
where $\diag\v h$ is a matrix with $\v h$ on the diagonal and zeroes
everywhere else; and $\Delta\p{\v v}$ is some constant that may depend
on $\v v$.  If the interaction Hamiltonian $H_0$ is translationally
invariant, then all $h_q=h$ and $\v h\oc\v v=h\v v$, so the condition
in \eqref{eq:cond_1} simplifies to
\begin{align}
  \m h \c\v v = \sp{\Delta\p{\v v}+h} \v v.
\end{align}
If we {\it construct} a perturbation $\V_1\p{\v v_\Delta}$ with
coefficients $\v v_\Delta$ that satisfy \eqref{eq:cond_1} with
eigenvalue $\Delta$, then
\begin{align}
  H_0 \V_1\p{\v v_\Delta} \ket\psi
  = \sp{E_0 + \Delta} \V_1\p{\v v_\Delta} \ket\psi.
  \label{eq:H_V1_psi}
\end{align}
Interestingly, the energy of the (unnormalized) state
$\V_1\p{\v v}\ket\psi$ depends only on the coefficients $\v v$, and is
entirely independent of the state $\ket\psi$ or choice of single-spin
trace-zero operator used to build $\V_1$ in \eqref{eq:perturbations}.
Finding operators $\V_1\p{\v v}$ that generate eigenvectors of the
interaction Hamiltonian $H_0$ when they act on fully symmetric states
$\ket\psi$ thus reduces to finding eigenvectors $\v v$ of the
interaction matrix $\m h-\diag\v h$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Two-body product}
\label{sec:PXYP}
